---
title: "Describing Data"
date: "1/27/2022"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mosaic)
```

Calculating treatment means for **one factor**:

```{r eval = FALSE, echo = TRUE}
library(mosaic)
favstats(Y~X, data=YourDataSet)
```

Example code:

<a href="javascript:showhide('mosaic_mean')">

::: hoverchunk
[ library(mosaic) ]{.tooltipr}[mosaic is an R Package that is useful in
the teaching of statistics to beginning programmers.]{.tooltiprtext}
<br>[ library(pander) ]{.tooltipr}[pander is an R Package that makes R
output look pretty]{.tooltiprtext} <br>[ favstats( ]{.tooltipr}[a
function from the mosaic package that returns a set of favorite summary
statistics]{.tooltipRtext} [ Temp ]{.tooltipr}[This is our response
variable. From `?airquality` you can see in the help file that Temp is
the maximum daily temp in degrees F at La Gaurdia Aiport during
1973]{.tooltipRtext} [ \~ ]{.tooltipr}["\~" is the tilde symbol. It can
be interpreted as "y broken down by x"; "y modeled by x"; "y explained
by x", etc. Where y is on the left of the tilde and x is on the right.
]{.tooltipRtext} [ Month, ]{.tooltipr}["Month" is a column from the
airquality dataset that can be treated as qualitative.]{.tooltiprtext} [
data = airquality ]{.tooltipr}[You have to tell R what dataset the
variables Temp and Month come from. 'airquality' is a preloaded dataset
in R. ]{.tooltiprtext} [ ) ]{.tooltipr}[Functions must always end with a
closing parenthesis.]{.tooltiprtext} [  Click to view output 
]{.tooltipr style="float:right;"}[Click to View Output.]{.tooltiprtext}
:::

</a>

::: {#mosaic_mean style="display:none;"}
```{r, echo=FALSE}
favstats(Temp~Month, data=airquality)
```
:::
