---
title: "BF1 R Instructions"
output: html_document
---

```{r message = FALSE, include = FALSE}
library(tidyverse)
```

<a href="javascript:showhide('bff1')">
<div class="hoverchunk">
  <span class="tooltipr">
df
  <span class="tooltiprtext">A name you come up with for your dataset</span>
  </span><span class="tooltipr">
<-
  <span class="tooltiprtext">The assignment operator. The result to the right of it gets stored in an object specified on the left</span>
  </span><span class="tooltipr">
read_csv(
  <span class="tooltiprtext">a command from the tidyverse to read in csv files</span>
  </span><span class="tooltipr">
"data/toothpaste_BF2.csv"
  <span class="tooltiprtext">The path to the csv file containing the data</span>
  </span><span class="tooltipr">
)
  <span class="tooltiprtext">Functions always end with a closing parenthesis</span>
  </span><br/><span class="tooltipr">
plaque_aov<span class="tooltiprtext">A name you come up with for your model</span>
  </span><span class="tooltipr">
<-
    <span class="tooltiprtext">The assignment operator. The result to the right of it gets stored in an object specified on the left</span>
  </span><span class="tooltipr">
aov(
  <span class="tooltiprtext">A function to define the model</span>
  </span><span class="tooltipr">
Plaque
  <span class="tooltiprtext">The response, or y, variable in the model. It is numeric.</span>
  </span><span class="tooltipr">
~
  <span class="tooltiprtext">The ~ is like an equal sign in the model. Items on the left of ~ represent y, on the right you define independent factors (i.e. x's).</span>
  </span><span class="tooltipr">
Brush,
  <span class="tooltiprtext">The independent variable containing the names for the 4 types of toothbrushes.</span>
  </span><span class="tooltipr">
data = df
  <span class="tooltiprtext">Tell the model to look in the dataset named "df" for Plaque and Brush variables</span>
  </span><span class="tooltipr">
)
<span class="tooltiprtext">Functions always end with a closing parenthesis</span>
  </span><br/><span class="tooltipr">
summary(
  <span class="tooltiprtext">Give important information about an object. When called on an aov object the default is to print the ANOVA table</span>
  </span><span class="tooltipr">
plaque_aov
  <span class="tooltiprtext">
  Whatever you named your ANOVA model in the previous line
  </span></span>
  <span class="tooltipr">
)
  <span class="tooltiprtext">
  Functions always end with a closing parenthesis
  </span></span><br/><span class="tooltipr" style="float:right;">
Click to view output 
  <span class="tooltiprtext">Click to View Output.</span>
</span>
</div>
</a>
<div id="bff1" style="display:show;">

```{r message = FALSE, echo = FALSE}
df <- read_csv(here::here("data/toothpaste_BF2.csv"))
plaque_aov <- aov(Plaque ~ Brush, data = df)
summary(plaque_aov)
```

</div>