---
output: html_document
---

<script type="text/javascript">
 function showhide(id) {
    var e = document.getElementById(id);
    e.style.display = (e.style.display == 'block') ? 'none' : 'block';
 }
 
 function openTab(evt, tabName) {
    var i, tabcontent, tablinks;
    tabcontent = document.getElementsByClassName("tabcontent");
    for (i = 0; i < tabcontent.length; i++) {
        tabcontent[i].style.display = "none";
    }
    tablinks = document.getElementsByClassName("tablinks");
    for (i = 0; i < tablinks.length; i++) {
        tablinks[i].className = tablinks[i].className.replace(" active", "");
    }
    document.getElementById(tabName).style.display = "block";
    evt.currentTarget.className += " active";
}
 
</script>

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(emmeans)
library(tidyverse)
library(pander)

```

**Caution**, the `contrasts()` function from the `stats` package in base R will produce the correct p-value for the test of a contrast, but without extra work will not produce the correct estimate of the contrast itself. For this reason, we illustrate estimating and testing the contrast with the `emmeans` package, which stands for "estimated marginal means".

The first step is to create the model. Then use the `emmeans()` command to create a grid of factor level summary statistics, including: means, standard deviations, standard error, degrees of freedom associated with the standard error estimate, and confidence intervals around the mean. Unlike a `summarize()` or `favstats()` command, `emmeans()` has the output structured so that it can easily be used in the next step. Store the grid of means into a new object.

```{r eval = FALSE, echo = TRUE}
myaov <- aov(Y ~ X, data = df)
mymeans <- emmeans(myaov, "X")

```

-   `myaov` is some name you come up with to store the results of the `aov()` model.
-   `Y` must be a "numeric" vector of the quantitative response variable.
-   `X` is a qualitative variable (should have class(X) equal to factor or character. If it does not, use `factor(X)` inside the `aov(Y ~ factor(X),...)` command.
-   `df` is the name of your data set.
-   `mymeans` is some name you come up with to store the results of the `emmeans()` command.


Define the contrasts you are interested in testing inside the `contrast()` function, which returns the hypothesis test results. You can also feed the result into a `confint()` function if you prefer confidence intervals over p-values.

```{r eval = FALSE, echo = TRUE}
contrast(mymeans,list(name_of_contrast1 = coefficient vector,
                       name_of_contrast2 = another coefficient vector))) 
```

`name_of_contrast` are descriptive names you should give to the contrast to help you remember what it represents. The `coefficient vector` is how you define the contrast.

We will repeat the contrasts we did by hand in the sections above, but this time using R.

**Example Code Using Toothbrush Experiment:**

<a href="javascript:showhide('contrast_brush')">
<div class="hoverchunk">
<span class="tooltipr">
df
  <span class="tooltiprtext">The name you want for your dataset</span>
  </span><span class="tooltipr">
<-
    <span class="tooltiprtext">The assignment operator. The result to the right of it gets stored in an object specified on the left</span>
  </span><span class="tooltipr">
read_csv("../data/toothpaste_BF2.csv")
  <span class="tooltiprtext">A tidyverse command to read the data in from the specified path</span>
  </span><br/><span class="tooltipr">
plaque_aov
  <span class="tooltiprtext">Name you want for your ANOVA model</span>
  </span><span class="tooltipr">
<-
    <span class="tooltiprtext">The assignment operator. The result to the right of it gets stored in an object specified on the left</span>
  </span><span class="tooltipr">
aov(
  <span class="tooltiprtext">A function to define the model</span>
  </span><span class="tooltipr">
Plaque
  <span class="tooltiprtext">The response, or y, variable in the model. It is numeric.</span>
  </span><span class="tooltipr">
~
  <span class="tooltiprtext">The ~ is like an equal sign in the model. Items on the left of ~ represent y, on the right you define independent factors (i.e. x's).</span>
  </span><span class="tooltipr">
Brush,
  <span class="tooltiprtext">The independent variable containing the names for the 4 types of toothbrushes.</span>
  </span><span class="tooltipr">
data = df
  <span class="tooltiprtext">Tell the model to look in the dataset named "df" for Plaque and Brush variables</span>
  </span><span class="tooltipr">
)
<span class="tooltiprtext">Functions always end with a closing parenthesis</span>
  </span><br/><span class="tooltipr">
brush_means
  <span class="tooltiprtext">The name you want for the output of the emmeans command</span>
  </span><span class="tooltipr">
<-
    <span class="tooltiprtext">The assignment operator. The result to the right of it gets stored in an object specified on the left</span>
  </span></span>
  <span class="tooltipr">
emmeans(
  <span class="tooltiprtext">
  Function to calculate stats about marginal means
  </span></span><span class="tooltipr">
plaque_aov,
  <span class="tooltiprtext">aov model created in previous step</span>
  </span><span class="tooltipr">
"Brush"
  <span class="tooltiprtext">Factor for whose levels you want to calculate means</span>
  </span><span class="tooltipr">
)
<span class="tooltiprtext">Functions always end with a closing parenthesis</span>
  </span><br/><span class="tooltipr">
contrast_results
<span class="tooltiprtext">Name you want to store contrast results in</span>
  </span><span class="tooltipr">
<-
    <span class="tooltiprtext">The assignment operator. The result to the right of it gets stored in an object specified on the left</span>
  </span><span class="tooltipr">
contrast(
    <span class="tooltiprtext">Function to define and test contrasts</span>
  </span><span class="tooltipr">
brush_means,
    <span class="tooltiprtext">Grid of stats about marginal means you named in the previous step</span></span><span class="tooltipr">
list(
    <span class="tooltiprtext">create a list object, which allows you to pass multiple contrast coefficient vectors</span>
  </span><span class="tooltipr">
man_v_osc =
    <span class="tooltiprtext">A descriptive name to help you remember what the contrast represents</span>
  </span><span class="tooltipr">
c(1,-1,0,0)
    <span class="tooltiprtext">Vector of coefficients used to define the contrast</span>
  </span><span class="tooltipr">
,
    <span class="tooltiprtext">Seperator to allow additional inputs to the list</span>
  </span><br/><span class="tooltipr">
&nbsp;&nbsp;&nbsp;man_v_others = 
    <span class="tooltiprtext">A descriptive name to help you remember what the contrast represents</span>
  </span><span class="tooltipr">
c(1,-(1/3),-(1/3),-(1/3))
    <span class="tooltiprtext">Vector of coefficients used to define the contrast</span>
  </span><span class="tooltipr">
),
    <span class="tooltiprtext">A list is closed with a parenthesis</span>
  </span><span class="tooltipr">
adjust = 
    <span class="tooltiprtext">Specify what type of adjustment (if any) to make for multiple testing. Default is "none" if this argument is not included.</span>
  </span><span class="tooltipr">
"none"
    <span class="tooltiprtext">Read help at ?summary.emmGrid for other acceptable values</span>
  </span><span class="tooltipr">
)
<span class="tooltiprtext">Functions always end with a closing parenthesis</span>
  </span><br/><span class="tooltipr">
contrast_results
    <span class="tooltiprtext">View the test results stored in this object in the previous step</span>
  </span><br/><span class="tooltipr">
confint(
  <span class="tooltiprtext">Function to create confidence intervals around contrasts</span>
  </span><span class="tooltipr">
contrast_results
    <span class="tooltiprtext">Name of object where you stored contrasts</span>
  </span><span class="tooltipr">
)
<span class="tooltiprtext">Functions always end with a closing parenthesis</span>
  </span><br/><span class="tooltipr" style="float:right;">
Toggle output 
  <span class="tooltiprtext">Toggle Output.</span>
</span>
</div>
</a>
<div id="contrast_brush" style="display:show;">

```{r message = FALSE, echo = FALSE}
df <- read_csv(here::here("data/toothpaste_BF2.csv"))
plaque_aov <- aov(Plaque ~ Brush, data = df)
brush_means <- emmeans(plaque_aov, "Brush")
contrast_results <- contrast(brush_means,list(man_v_osc = c(1,-1,0,0),
                      man_v_others = c(1,-(1/3),-(1/3),-(1/3))), adjust="none")
contrast_results
```

```{r message = FALSE, echo = FALSE}
confint(contrast_results)
```

</div>
